### 2015-03-16 doc/thesis/system.tex
Rearranging the explanation of typing rules

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

The rule \textsc{T-ASSIGNMENT} is the general rule,
\textsc{T-EXPLIST1} is the rule that type checks an expression list
where all expressions can only produce a single value, and
\textsc{T-LHSLIST} is the rule that type checks a list of left-hand values.

Note that our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.

@@

Back to our previous example, it type checks through these three rules
that we just mentioned because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
\]

As another example, lets assume that $x$, $y$, and $z$ are variables in
the environment with types $\Integer$, $\String$, and $\String \cup \Nil$,
the assignment
\[
x, y, z = 1, ``foo"
\]
type checks because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

@@

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that produce only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced.
We can define \emph{first} inductively as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(T{*}) & = T \cup \Nil\\
first(T \times P) & = T\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

As an example, assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

Our type system also catches arity mismatch.
For instance, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, assuming that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
the function call
\[
f(1)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
f(1,2,3)
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

@@

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\overline{p}\;\}$ and $\{\;\overline{p},me\;\}$.
The first uses a list of table fields $p_{1}, ..., p_{n}$,
where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
The second uses a list of table fields and an expression that can
produce multiple values.
We did not include in the abstract syntax a table constructor that uses
lists of expressions, because we can express this form using lists of
table fields.
For instance, we can use the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{p}\;|}
      {\env_{1}, \penv \vdash \{\;\overline{p}\;\}:T, \env_{n+1}}
\end{array}
\]

As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$.

First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
field $p_{i}$.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check a table field $p$
producing a pair $(K,V)$ and a new type environment $\env_{2}$. 
In a pair $(K,V)$, $K$ is the key type, while $V$ is the value type.

Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}
does not type check,
as aliasing $a$ produces the type $\{``foo":\Integer\}_{closed}$
that is not a subtype of $\{``foo":\Integer \cup \Nil\}_{closed}$,
the type of $b$.
Our type system has this behavior to warn programmers about
potential unsafe behaviors after this kind of alias.
In this example, it is unsafe to assign $\mathbf{nil}$ to the field
$``foo"$ of $b$, as it changes the value that is stored in the field
$``foo"$ of $a$.

As another example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
that is not a subtype of $\{\}_{open}$, the type of $b$.
In this example, it is unsafe to add the field $``x"$ to $b$,
as it changes the value that is stored in the field $``x"$ of $a$.

Aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
For this reason, the rule \textsc{T-IDREAD1} changes the type of
the original reference from \emph{unique} to \emph{open}.

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as defined by
rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
\end{array}
\]

@@

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b = a;$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check because we cannot add the field $``x"$ to $b$,
as its type is \emph{closed} through \textsc{T-IDWRITE}.
Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
$\emph{open}$, and that is the reason why we can add the field
$``x"$ to the type of $a$.
Aliasing $a$ also produces the type $\{\}_{closed}$, which is
the same type that $b$ has in left-hand side of the assignment.
After the assignment, the type of $b$ is \emph{closed} and thus
does not allow changing the value that is stored in the field
$``x"$ of $a$.

@@

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression that
obeys the \textsc{T-COERCE} rule below:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(id) \subtype T \;\;\;
       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
\end{array}
\]

@@

The rule \textsc{T-COERCE} allows our type system to type check the following example:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
The rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
\end{array}
\]

Rule \textsc{T-FUNCTION1} uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original types.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original types.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

>>>>

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$.
Let us see how our type system type checks the following assignment
\[
x, y = 1, ``foo"
\]

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} which is the rule
that type checks an expression list where all expressions can
only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

@@

Then, our type system type checks the list of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} which is the rule that type
checks a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

After type checking the right-hand side and the left-hand side of an assignment,
our type system checks whether their types are consistent.
The rule \textsc{T-ASSIGNMENT} is the general rule that express this idea:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\]

Back to our example, it type checks through rule \textsc{T-ASSIGNMENT} because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}
\]

As another example, lets assume that $x$, $y$, and $z$ are variables in
the environment with types $\Integer$, $\String$, and $\String \cup \Nil$.
The assignment
\[
x, y, z = 1, ``foo"
\]
type checks because
\[
1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times (\String \cup \Nil) \times \Value{*}
\]

@@

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

We can define \emph{first} inductively as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(T{*}) & = T \cup \Nil\\
first(T \times P) & = T\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

As an example, let us assume that $f$ is a local function in the environment,
and that $f$ has type $\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*} \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Value{*}
\]

Our type system also catches arity mismatch.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, let us assume that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$.
The function call
\[
f(1)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
f(1,2,3)
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

@@

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\overline{[e_{1}] = e_{2}}\;\}$ and
$\{\;\overline{[e_{1}] = e_{2}},me\;\}$.
The first uses a list of table fields $([e_{1}] = e_{2})_{1}, ..., ([e_{1}] = e_{2})_{n}$.
The second uses a list of table fields and an expression that can
produce multiple values.

@@

As a more interesting example, let us see how our type system type checks
the table constructor $\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$.

First, our type system uses the auxiliary relation
$\env_{1}, \penv \vdash [e_{1}] = e_{2} : (K,V), \env_{2}$ to type check each
table field.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, checking a table field $[e_{1}] = e_{2}$
produces a pair $(K,V)$ and a new type environment $\env_{2}$.
A pair $(K,V)$ means that $e_{1}$ has type $K$ and $e_{2}$ has type $V$,
where $K$ is the type of the key and $V$ is the type of the value.

After type checking each table field, our type system uses each pair $(K,V)$
to build the table type that express the type of a given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

Well-formed table types also do not allow \emph{unique} and
\emph{open} table types to appear in the type of the values.
We made this restriction because our type system does not keep
track of aliases to table fields.
This means that allowing \emph{unique} and \emph{open} table
types to appear in the type of a value would allow the
creation of unsafe aliases.
The rules that type check table fields use the auxiliary function
\emph{close} to close the type of the values in a table type.

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{i}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;|
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{n+1}}
\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.

As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check, as aliasing $a$ produces the type $\{\}_{closed}$
that is not a subtype of $\{\}_{open}$, the type of $b$.
Our type system has this behavior to warn programmers about
potential unsafe behaviors after this kind of alias.
In this example, it is unsafe to add the field $``x"$ to $b$,
as it changes the value that is stored in the field $``x"$ of $a$.

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as leaving them
\emph{unique} and \emph{open} would allow the creation of
unsafe references.

As an example,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b = a;$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check because we cannot add the field $``x"$ to $b$,
as its type is \emph{closed}.
Aliasing $a$ changes the type of $a$ from $\emph{unique}$ to
$\emph{open}$, and that is the reason why we can add the field
$``x"$ to the type of $a$.
Aliasing $a$ also produces the type $\{\}_{closed}$, which is
the same type that $b$ has in left-hand side of the assignment.
After the assignment, the type of $b$ is \emph{closed} and thus
does not allow changing the value that is stored in the field
$``x"$ of $a$.

Rule \textsc{T-IDWRITE} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env_{1}, \penv \vdash id_{l}:close(T_{2}), \env_{1}[id \mapsto close(T_{2})]}
\end{array}
\]

@@

@@

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression ${<}T{>} \; id$.
For instance, we can use this type coercion expression to make the following
example type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{ \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& \multicolumn{2}{l}{$a[``y"] \; {<}\String{>} = ``bar";$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String, ``y":\String \cup \Nil \}_{closed} =$}\\
& & \multicolumn{1}{l}{${<}\{``x":\String, ``y":\String \cup \Nil\}_{open}{>} \; a \; \mathbf{in} \; a[``z"] \; {<}\Integer{>} = 1$}
\end{tabular}
\end{center}

Rule \textsc{T-COERCE} defines the behavior of the type coercion expression:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(id) \subtype T \;\;\;
       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
\end{array}
\]

@@

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
To do that, our type system uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original types.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original types.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{crall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{crset}.
Then, the function \emph{rv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

Rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
\end{array}
\]
